Bioconductor workflow for single-cell RNA sequencing: Dimensionality reduction, clustering, and lineage inference

### 2017-06-28 workflow/workflow.Rmd
add new slingshot and gam code from kelly

#bioconductor:
library(clusterExperiment)
library(BiocParallel)
library(zinbwave)
#github
library(scone)
library(slingshot)
#CRAN
library(doParallel)
library(RColorBrewer)
set.seed(20)
if(packageVersion("zinbwave")<'0.99.4.2') stop("must have current develop version to avoid bugs")
```

@@

```{r scone, fig.cap="Plot generated by scone to filter low quality cells."}
# Metric-based Filtering sample-filtering
data("housekeeping")
hk = rownames(se)[toupper(rownames(se)) %in% housekeeping$V1]

@@

```{r sconeFilt}
# Simplify to a single logical
mfilt <- !apply(simplify2array(mfilt[!is.na(mfilt)]), 1, any)
se <- se[, mfilt]
dim(se)
```

@@

```{r}
# Filtering to top 1000 most variable
vars <- rowVars(log1p(assay(se)))
names(vars) <- rownames(se)
vars <- sort(vars, decreasing = TRUE)
core <- se[names(vars)[1:1000],]
save(core, file="../data/oe_se_1000Var.rda")
```

@@

We now want to demonstrate how to use the R software package `slingshot` to infer branching lineages and order cells by developmental progression. We connect the clusters identified by `RSEC` with a minimum spanning tree to learn the global lineage structure. Then, we refine this structure and order cells using highly stable simultaneous principal curves to infer smooth, branching lineages.

@@

Kelly, would you like to give details about slingshot algorithm?

@@

lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1")
curves <- get_curves(X, clus.labels = cl, lineages = lineages)
```

```{r curves, fig.cap="Cells colored by cluster in a 4-dimensional space with smooth curves representing each inferred lineage."}
plot_curves(X, cl, curves, col.clus = pal)
```

```{r tree, fig.cap="Cells colored by cluster in a 4-dimensional space with connecting lines between cluster centers representing the inferred structure."}
plot_tree(X, cl, lineages, col.clus = pal)
```

```{r lineages}
print(lineages$lineage1)
print(lineages$lineage2)
print(lineages$lineage3)
```

In the workflow, we recover a reasonable pseudotime ordering of the clusters using the unsupervised version of slingshot. However, in some other cases, we have noticed that we need to give more guidance to the algorithm to find the correct ordering. Slingshot has the option for the user to provide the end cluster(s). Here is the code to use slingshot in a supervised setting

```{r slingshotsupervised, eval=FALSE}
lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1",
                         end.clus = c("c3", "c6"))
curves <- get_curves(X, clus.labels = cl, lineages = lineages)
plot_curves(X, cl, curves, col.clus = pal)
plot_tree(X, cl, lineages, col.clus = pal)

print(lineages$lineage1)
print(lineages$lineage2)
print(lineages$lineage3)
```

@@

Kelly, would you like to complete this part? Here is the kind of plots we would want to present.

```{r de, eval = FALSE}
de <- read.csv('../data/oe_markers.txt', stringsAsFactors = F, header = F)
de <- de$V1
plotHeatmap(ceObj, 
            visualizeData = assays(se)$normalizedValues[rownames(se) %in% de, ],
            clusterSamplesData = "dendrogramValue",
            whichClusters = "primary",
            main = 'Normalized values, 1000 most variable genes',
            breaks = 0.99)
```

>>>>

#bioconductor:
library(clusterExperiment)
library(BiocParallel)
library(zinbwave)
#github
library(slingshot)
#library(scone)
#CRAN
library(doParallel)
library(RColorBrewer)
library(gam)
set.seed(20)
if(packageVersion("zinbwave")<'0.99.4.2') stop("must have current develop version to avoid bugs")
```

@@

```{r scone, fig.cap="Plot generated by scone to filter low quality cells.", eval=FALSE}
# Metric-based Filtering sample-filtering
data("housekeeping")
hk = rownames(se)[toupper(rownames(se)) %in% housekeeping$V1]

@@

```{r sconeFilt, eval=FALSE}
# Simplify to a single logical
mfilt <- !apply(simplify2array(mfilt[!is.na(mfilt)]), 1, any)
se <- se[, mfilt]
dim(se)
```

@@

```{r,eval=FALSE}
# Filtering to top 1000 most variable
vars <- rowVars(log1p(assay(se)))
names(vars) <- rownames(se)
vars <- sort(vars, decreasing = TRUE)
core <- se[names(vars)[1:1000],]
```

```{r}
load("../data/oe_se_1000Var.rda")
```

@@

We now want to demonstrate how to use the R software package `slingshot` to infer branching lineages and order cells by developmental progression. We connect the clusters identified by `RSEC` with a minimum spanning tree (MST) to learn the global lineage structure. Then, we refine this structure and order cells using highly stable simultaneous principal curves to infer smooth, branching lineages.

@@

The two steps of the slingshot algorithm are implemented in the functions 'getLineages' and 'getCurves'. The first takes the low-dimensional representation of the cells and the vector of cluster labels identified above. It fits an MST to the clusters and identifies lineages as paths through this tree. The output of 'getLineages' is a 'SlingshotDataSet' object containing all the information used to fit the tree and identify lineages. 'getCurves' takes this object as input and fits simultaneous principal curves to the identified lineages. These functions can be run separately, as below, or jointly by the wrapper function, 'slingshot'.

@@

lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1")
lineages <- getCurves(lineages)
```

```{r curves, fig.cap="Cells colored by cluster in a 4-dimensional space with smooth curves representing each inferred lineage."}
pairs(lineages, type='curves', pal[primaryClusterNamed(ceObj)])
```

```{r tree, fig.cap="Cells colored by cluster in a 4-dimensional space with connecting lines between cluster centers representing the inferred structure."}
pairs(lineages, type='lineages', col = pal[primaryClusterNamed(ceObj)])
```

```{r lineages}
lineages
```

In the workflow, we recover a reasonable pseudotime ordering of the clusters using the unsupervised version of slingshot. However, in some other cases, we have noticed that we need to give more guidance to the algorithm to find the correct ordering. 'getLineages' has the option for the user to provide known end cluster(s). Here is the code to use slingshot in a supervised setting:

```{r slingshotsupervised, eval=FALSE}
lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1",
                         end.clus = c("c3", "c6"))
lineagees <- getCurves(lineages)
pairs(lineages, type='curves', pal[primaryClusterNamed(ceObj)])
pairs(lineages, type='lineages', col = pal[primaryClusterNamed(ceObj)])

lineages
```

@@

After ordering the cells, we are interested in finding genes that have non-constant expression patterns over pseudotime. We can see global trends in expression by subsetting our original data to the most variable genes and plotting their expression values in a heatmap, ordered by pseudotime.

More formally, we are interested in testing for variable expression over pseudotime. A general additive model (GAM) is a flexible, nonlinear way to test the relationship between a gene's normalized expression values and pseudotime. We implemented this model for each gene and show the top 100 genes by p-value in the heatmap below.

```{r fitgam}
t <- pseudotime(lineages)[,1]
y <- assays(se)$normalizedValues[, our_cl != "-1"]
gam.pval <- apply(y,1,function(z){
  d <- data.frame(z=z, t=t)
  tmp <- gam(z ~ lo(t), data=d)
  p <- summary(tmp)[4][[1]][1,5]
  p
})
```

```{r heatmapsignificant}
topgenes <- names(sort(gam.pval, decreasing = FALSE))[1:100]
heatdata <- y[rownames(se) %in% topgenes, order(t, na.last = NA)]
heatclus <- cl[order(t, na.last = NA)]
ce <- clusterExperiment(heatdata, heatclus, transformation = identity)
plotHeatmap(ce, clusterSamplesData = "orderSamplesValue", breaks = .99)
```
